---
title: "Deterministic Sensitivity Analysis: Generation"
author: "Fernando Alarid-Escudero, Greg Knowlton, Eva Enns, and the DARTH Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Deterministic Sensitivity Analysis: Generation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```

# Overview

`dampack` has the functionality to interface with a user-defined decision model in `R` to conduct deterministic sensitivity analysis (DSA) on parameters of interest. DSA is a method for assessing how the results of a decision analytic model vary over a pre-specified range of values for a parameter of interest. In a DSA, the model is run for each value of the parameter of interest over the pre-specified range, while holding all other parameter values fixed. If one parameter is varied, this is called a one-way DSA. If two parameters are varied, it is called a two-way DSA. We typically do not vary more than two parameters at a time, as the output becomes difficult to visualize and interpret. A more comprehensive assessment of how model outptus depend on model parameters can be done through a probabilistic sensitivity analysis (PSA) (type `vignette("psa_generation", package = "dampack")` in the console after installing the `dampack` package to see a vignette describing how to use `dampack` for PSA).

A simple single-parameter DSA starts with choosing the model parameter to be investigated. Next, the modeler specifies a range for this parameter and the number of evenly spaced points along this range at which to evaluate model outcomes. The model is then run for each element of the vector of parameter values by setting the parameter of interest to the value, holding all other model parameters at their default base case values.

## Decision Model Format

`dampack` includes functionality to generate DSA results for any user-defined decision analytic model that is written in `R`. The user-defined model must be written as a function takes in a list of all model parameter values as its first argument and outputs a data frame of the model outcomes for each strategy evaluated by the model. The first column of the output data frame must be the strategy names (as strings) with any number of additional outcomes (costs, QALYs, but also additional epidemiological outcomes like infections averted) stored in subsequent columns. Thus, each row of the output data frame consists of the strategy's name followed by the corresponding outcome values for that strategy. The user-defined model function may have additional required or optional input arguments; values for these additional arguments will need to be passed to the function through the `dampack` DSA functions.

As an example, we define the function `run_decision_model` below, which is a four-state cohort state transition model (also known as a Markov model) that returns the total remaining life-years, quality-adjusted life-years, and costs incured by a cohort progressing through a disease known as the Sick-Sicker model. For a deeper discusison of state transition models and more complex variations of this state transition model, see Alarid-Escudero F, Krijkamp EM, Enns EA, Yang A, Hunink MGM, Pechlivanoglou P, Jalal H. Cohort state-transition models in R: A Tutorial. arXiv:200107824v1. 2020:1-31. 

```{r}
run_decision_model <- function(l_params_all, verbose = FALSE) {
  with(as.list(l_params_all), {
    # l_params_all must include:
    # -- transition probability parameters: p_HD, p_S1S2, hr_S1D, hr_S2D, 
    # -- initial cohort distribution: v_init
    # -- vector of state utilities: v_u = c(u_H, u_S1, u_S2, u_D)
    # -- vector of state costs: v_c = c(c_H, c_S1, c_S2, c_D)
    
    # compute internal parameters as a function of external parameters
    r_HD    = -log(1 - p_HD) # rate of death in healthy
    r_S1D   = hr_S1D * r_HD 	# rate of death in sick
    r_S2D   = hr_S2D * r_HD  	# rate of death in sicker
    p_S1D   = 1 - exp(-r_S1D) # probability to die in sick
    p_S2D   = 1 - exp(-r_S2D) # probability to die in sicker
    
    ####### INITIALIZATION ##########################################
    # create the cohort trace
    m_M <- matrix(NA, nrow = n_t + 1 , 
                  ncol = n_s,
                  dimnames = list(0:n_t, v_n)) # create Markov trace
    
    m_M[1, ] <- v_init # initialize Markov trace
    
    # initialize transition probability matrix
    m_P <- matrix(0,
                  nrow = n_s, 
                  ncol = n_s,
                  dimnames = list(v_n, v_n))
    # fill in the transition probability array
    ### From Healthy
    m_P["H", "H"]  <- 1 - (p_HS1 + p_HD)
    m_P["H", "S1"] <- p_HS1
    m_P["H", "D"]  <- p_HD
    ### From Sick
    m_P["S1", "H"]  <- p_S1H
    m_P["S1", "S1"] <- 1 - (p_S1H + p_S1S2 + p_S1D)
    m_P["S1", "S2"] <- p_S1S2
    m_P["S1", "D"]  <- p_S1D
    ### From Sicker
    m_P["S2", "S2"] <- 1 - p_S2D
    m_P["S2", "D"]  <- p_S2D
    ### From Dead
    m_P["D", "D"] <- 1
    
    # check rows add up to 1
    if (!isTRUE(all.equal(as.numeric(rowSums(m_P)), as.numeric(rep(1, n_s))))) {
      stop("This is not a valid transition Matrix")
    }
    
    ############# PROCESS ###########################################
    
    for (t in 1:n_t){ # throughout the number of cycles
      m_M[t + 1, ] <- m_M[t, ] %*% m_P # estimate the Markov trace for cycle the next cycle (t + 1)
    }
    
    ####### EPIDEMIOLOGICAL OUTPUT  ###########################################
    #### Overall Survival (OS) ####
    v_os <- 1 - m_M[, "D"]  # calculate the overall survival (OS) probability for no treatment
    
    #### Disease prevalence #####
    v_prev <- rowSums(m_M[, c("S1", "S2")])/v_os
    
    #### Proportion of sick in S1 state #####
    v_prop_S1 <- m_M[, "S1"] / v_prev
    
    ####### RETURN OUTPUT  ###########################################
    out <- list(m_M = m_M,
                m_P = m_P,
                Surv = v_os[-1],
                Prev = v_prev[-1],
                PropSick = v_prop_S1[c(11, 21, 31)])
    
    return(out)
  }
  )
}
```


that evaluates two strategies ("no treatment" and "treatment") for a hypothetical disease. `run_decision_model` has as inputs `l_params_all`, which is the input parameter values list. It also has an additional optional argument `wtp` which allows the user to change the willingness-to-pay used in the model when calculating the net monetary benefit (NMB) model outcome. 

```{r}
library(dampack)
v_names_str <- c("No Treatment", "Treatment")  
## Number of strategies
n_str <- length(v_names_str)
## Markov model parameters
n_age_init <- 25                # age at baseline
n_age_max  <- 55                # maximum age of follow up
n_t  <- n_age_max - n_age_init  # time horizon, number of cycles
v_n  <- c("H", "S1", "S2", "D") # the 4 states of the model: Healthy (H), Sick (S1), Sicker (S2), Dead (D)
n_s <- length(v_n)              # number of health states 

calculate_ce_out <- function(l_params_all, wtp = 100000){
  # User defined
  with(as.list(l_params_all), {
    ## Create discounting vectors
    v_dwc <- 1 / ((1 + d_e) ^ (0:(n_t))) # vector with discount weights for costs
    v_dwe <- 1 / ((1 + d_c) ^ (0:(n_t))) # vector with discount weights for QALYs
    
    ## Run STM model at a parameter set for each intervention
    l_model_out_no_trt <- run_decision_model(l_params_all = l_params_all)
    l_model_out_trt    <- run_decision_model(l_params_all = l_params_all)
    
    ## Cohort trace by treatment
    m_M_no_trt <- l_model_out_no_trt$m_M # No treatment
    m_M_trt    <- l_model_out_trt$m_M    # Treatment
    
    ## Vectors with costs and utilities by treatment
    v_u_no_trt <- c(u_H, u_S1, u_S2, u_D)
    v_u_trt    <- c(u_H, u_Trt, u_S2, u_D)
    
    v_c_no_trt <- c(c_H, c_S1, c_S2, c_D)
    v_c_trt    <- c(c_H, c_S1 + c_Trt, c_S2 + c_Trt, c_D)
    
    ## Mean Costs and QALYs for Treatment and NO Treatment
    v_tu_no_trt <- m_M_no_trt %*% v_u_no_trt
    v_tu_trt    <- m_M_trt %*% v_u_trt
    
    v_tc_no_trt <- m_M_no_trt %*% v_c_no_trt
    v_tc_trt    <- m_M_trt %*% v_c_trt
    
    ## Total discounted mean Costs and QALYs
    tu_d_no_trt <- t(v_tu_no_trt) %*% v_dwe 
    tu_d_trt    <- t(v_tu_trt) %*% v_dwe
    
    tc_d_no_trt <- t(v_tc_no_trt) %*% v_dwc
    tc_d_trt    <- t(v_tc_trt)    %*% v_dwc
    
    ## Vector with total discounted mean Costs and QALYs
    v_tc_d <- c(tc_d_no_trt, tc_d_trt)
    v_tu_d <- c(tu_d_no_trt, tu_d_trt)
    
    ## Vector with discounted net monetary benefits (NMB)
    v_nmb_d <- v_tu_d * wtp - v_tc_d
    
    ## Dataframe with discounted costs, effectiveness and NMB
    df_ce <- data.frame(Strategy = v_names_str,
                        Cost     = v_tc_d,
                        Effect   = v_tu_d,
                        NMB      = v_nmb_d)
    
    return(df_ce)
  }
  )
}

run_decision_model <- function(l_params_all, verbose = FALSE) {
  with(as.list(l_params_all), {
    # compute internal parameters as a function of external parameters
    r_HD    = - log(1 - p_HD) # rate of death in healthy
    r_S1D   = hr_S1 * r_HD 	  # rate of death in sick
    r_S2D   = hr_S2 * r_HD  	# rate of death in sicker
    p_S1D   = 1 - exp(-r_S1D) # probability to die in sick
    p_S2D   = 1 - exp(-r_S2D) # probability to die in sicker
    
    ####### INITIALIZATION ##########################################
    # create the cohort trace
    m_M <- matrix(NA, nrow = n_t + 1 , 
                  ncol = n_s,
                  dimnames = list(0:n_t, v_n)) # create Markov trace
    
    m_M[1, ] <- c(1, 0, 0, 0) # initialize Markov trace
    
    # create transition probability matrix for NO treatment
    m_P <- matrix(0,
                  nrow = n_s, 
                  ncol = n_s,
                  dimnames = list(v_n, v_n))
    # fill in the transition probability array
    ### From Healthy
    m_P["H", "H"]  <- 1 - (p_HS1 + p_HD)
    m_P["H", "S1"] <- p_HS1
    m_P["H", "D"]  <- p_HD
    ### From Sick
    m_P["S1", "H"]  <- p_S1H
    m_P["S1", "S1"] <- 1 - (p_S1H + p_S1S2 + p_S1D)
    m_P["S1", "S2"] <- p_S1S2
    m_P["S1", "D"]  <- p_S1D
    ### From Sicker
    m_P["S2", "S2"] <- 1 - p_S2D
    m_P["S2", "D"]  <- p_S2D
    ### From Dead
    m_P["D", "D"] <- 1
    
    # check rows add up to 1
    if (!isTRUE(all.equal(as.numeric(rowSums(m_P)), as.numeric(rep(1, n_s))))) {
      stop("This is not a valid transition Matrix")
    }
    
    ############# PROCESS ###########################################
    
    for (t in 1:n_t){ # throughout the number of cycles
      m_M[t + 1, ] <- m_M[t, ] %*% m_P # estimate the Markov trace for cycle the next cycle (t + 1)
    }
    
    ####### EPIDEMIOLOGICAL OUTPUT  ###########################################
    #### Overall Survival (OS) ####
    v_os <- 1 - m_M[, "D"]  # calculate the overall survival (OS) probability for no treatment
    
    #### Disease prevalence #####
    v_prev <- rowSums(m_M[, c("S1", "S2")])/v_os
    
    #### Proportion of sick in S1 state #####
    v_prop_S1 <- m_M[, "S1"] / v_prev
    
    ####### RETURN OUTPUT  ###########################################
    out <- list(m_M = m_M,
                m_P = m_P,
                Surv = v_os[-1],
                Prev = v_prev[-1],
                PropSick = v_prop_S1[c(11, 21, 31)])
    
    return(out)
  }
  )
}
```

## Generating DSA results

`dampack` uses the function `run_owsa_det` to execute a one-way DSA and the function `run_twsa_det` to execute a two-way DSA.

*** To be deleted / incorporated above ***

and the function `run_twsa_det` to execute a two-way DSA. `run_owsa_det` and `run_twsa_det` generate the vector of parameter values for which the user-defined model should be run and then also runs the model.  

containing a list of the parameters from the `params_basecase` argument. In the example model shown below, this list is named `l_params_all`, and the variables contained in this list are the only variables that are allowed to be varied through the `params_range` argument in the DSA. Optionally, additional function inputs for `FUN` can be supplied through the `...` argument of `run_owsa_det`/`run_twsa_det`, but these inputs are not allowed to vary in the sensitivity analysis. These additional inputs must be arguments of `FUN`, like `wtp` in the example of `calculate_ce_out()` below. `FUN` and its component functions are also able to incorporate variables stored in the global environment, such as `n_age_init` or `n_age_max` in the example.

The user-defined model function must return a data.frame where the first column contains a character vector of the strategy names, and the subsequent columns contain numeric vectors of all relevant model outcomes. Each row of the data.frame will consist of a strategy name followed by the corresponding outcome values for that strategy. These model outcomes must be calculated internally within `FUN`. The model outcomes are not limited to typical outcomes like cost or effectiveness and can be any numerical outcome that the user chooses to model.

`dampack` uses the function `run_owsa_det` to execute one-way DSAs and the function `run_twsa_det` to execute two-way DSAs. In order to perform a DSA in `dampack`, the user must input the code for the decision analytic model in a standardized format that is compatible with `run_owsa_det`, and `run_twsa_det`. This is the same format required for the `FUN` argument of `run_psa`.

*** END DELETIONS ***



## DSA Parameter Format

The `params_range` input tells `run_owsa_det` or `run_twsa_det` which parameters will be varied in the DSA and over what ranges. This input must be a data.frame with 3 columns in the following order: "pars", "min", and "max". The number of samples from this range is determined by the input `nsamp`. The "pars" column is a character vector with the parameters of interest and must be a subset of the parameters from `params_basecase`.

when using `run_owsa_det`, supplying multiple parameters in `params_range` will instruct the function to execute a series of separate one-way deterministic sensitivity analyses and compile the results into a single `owsa` object for each requested `outcome`. When using `run_twsa_det`, exactly two parameters must be supplied in the `pars` column of the `params_range` data.frame argument.

```{r}
my_params_basecase <- list(p_HS1 = 0.15,
                           p_S1H = 0.5,
                           p_S1S2 = 0.105,
                           p_HD = 0.005, 
                           hr_S1 = 3, 
                           hr_S2 = 10, 
                           c_H = 2000,
                           c_S1 = 4000,
                           c_S2 = 15000, 
                           c_Trt = 12000,
                           c_D = 0,
                           u_H = 1,
                           u_S1 = 0.75,
                           u_S2 = 0.5,
                           u_D = 0,
                           u_Trt = 0.95,
                           d_e = 0.03,
                           d_c = 0.03)

my_params_range <- data.frame(pars = c("u_Trt", "c_Trt"),
                              min = c(0.9, 12000),
                              max = c(1, 24000))
```

## One-way Deterministic Sensitivity Analysis

In `run_owsa_det`, the user is free to specify any number of outcomes that are defined in the user-defined function (the `FUN` argument, which is `calculate_ce_out` in this example). If more than one outcome is provided, `run_owsa_det` will return a list containing one `owsa` object for each outcome. If the decision model in `FUN` is computationally slow and `nsamp` is large, `run_owsa_det` could take an extremely long time to run. Under these circumstances, it is recommended that you set the `progress` argument to `TRUE` in order to print a progress bar in the console while the function is running.

Each `owsa` object returned by `run_owsa_det` is a `data.frame` with four columns, `parameter`, `strategy`, `param_val`, and `outcome_val`. Each row's value in the `parameter` column denotes which parameter was swapped into  the basecase parameters, and `param_val` indicates which particular value of this parameter was used in that iteration of the decision model. The `outcome_val`s corresponding to each combination of `strategy`, `parameter`, and `param_val` are provided in long format within the `owsa` object.

To access a single `owsa` object associated with the `"NMB"` outcome within the returned list, we can simply write `owsa_det$owsa_NMB`.

```{r}
owsa_det <- run_owsa_det(params_range = my_params_range,
                         params_basecase = my_params_basecase,
                         nsamp = 100,
                         FUN = calculate_ce_out,
                         outcomes = c("Cost", "Effect", "NMB"),
                         strategies = c("No_Treatment", "Treatment"),
                         progress = FALSE)
```

## Two-way Deterministic Sensitivty Analysis

A two-way sensitivity analysis is used to assess how model outcomes vary over pre-specified ranges of two model parameters jointly. The general format of the function arguments for `dampack`'s two-way DSA function, `run_twsa_det`, are the same as those for `run_owsa_det`. In `run_twsa_det`, equally spaced sequences of length `nsamp` are created for the two parameters based on the inputs provided in the `params_range` argument. These two sequences of parameter values define an `nsamp` by `nsamp` grid over which `FUN` is applied to produce outcomes for every combination of the two parameters.

If more than one outcome is provided, `run_twsa_det` will return a list containing one `twsa` object for each outcome. To access a single `twsa` object associated with the `"NMB"` outcome within the returned list, we write `twsa_det$twsa_NMB`. Each `twsa` object is a long form `data.frame` that reports the outcome value for each each strategy at every parameter combination. The first two columns are named according to the two parameters specified in `params_range`. 

```{r}
twsa_det <- run_twsa_det(params_range = my_params_range,
                         params_basecase = my_params_basecase,
                         nsamp = 100,
                         FUN = calculate_ce_out,
                         outcomes = c("Cost", "Effect", "NMB"),
                         strategies = c("No_Treatment", "Treatment"), 
                         progress = FALSE)
```
