---
title: "Probabilistic Sensitivity Analysis: Basic Usage"
author: "Fernando Alarid-Escudero, Caleb Easterly, Eva Enns, and the DARTH Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Probabilistic Sensitivity Analysis: Basic Usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.align = "center"
)
```

The probabilistic sensitivity analysis object is a key part of the `dampack` package. In this vignette, we discuss the goals of a probabilistic sensitivity analysis, how to create the `dampack` object from an existing PSA, and what can be done once the PSA object is created in `dampack`.

## Probability Sensitivity Analysis: An Introduction

**theoretical background and stuff here**

## PSA in dampack

First, we need to load the `dampack` package.
```{r}
library(dampack)
```

### Read in data

**explanation of data here?**

The PSA object is created using the function `make_psa_obj`. We can see the types of data we need by examining what arguments this function takes.

```{r}
args(make_psa_obj)
```

To see more information, you can write `?make_psa_obj` in the R console. 

As the arguments (and the help text) show, we need the `cost`, `effectiveness`, `strategies` (optional), and `currency` (also optional). 

There is example data provided with dampack that allows us to illustrate the required structure of each of these parts of a PSA.

```{r}
data("example_psa")
```

We can see the structure of the object with `str()`:
```{r}
str(example_psa)
```

We can ignore the `wtp` part for now. The other parts of the `example_psa` object are:

- `strategies`: a list of three strategy names
```{r}
example_psa$strategies
```

- `cost`: a data frame of the cost of carrying out each strategy. The columns are the strategies and the
rows are the PSA samples (of which there were 10,000).
```{r}
head(example_psa$cost)
```

Note that the columns have names that suggest the strategy and the units. This is, however, not necessary. It is *critical* that the vector of strategies (`example_psa$strategies`) and the columns of the `cost` and `effectiveness` dataframes be in the same order, though.

- `effectiveness`: a data frame of the effects of carrying out each strategy, with the same structure as `costs`
```{r}
head(example_psa$effectiveness)
```

Now, let's make the PSA object:

```{r}
psa_obj <- make_psa_obj(cost = example_psa$cost,
                        effectiveness = example_psa$effectiveness,
                        parameters = example_psa$parameters,
                        strategies = example_psa$strategies,
                        currency = "$")
str(psa_obj)
```

### Scatterplot

**explanation here**

```{r fig.width = 7, fig.height = 5, fig.align = "center"}
plot(psa_obj)
```

### Cost-effectiveness acceptability curve

**explanation here**

```{r}
ceac_obj <- ceac(wtp = example_psa$wtp, psa = psa_obj)
head(ceac_obj)
```

We can get the summary of the cost-effective strategy in each range of willingness-to-pay:

```{r}
summary(ceac_obj)
```

We can also plot the cost-effectiveness acceptability curve.

```{r}
plot(ceac_obj, frontier = TRUE)
plot(ceac_obj, frontier = FALSE)
```


### Expected Value of Perfect Information

**explanation here**

```{r}
evpi <- calc_evpi(wtp = example_psa$wtp, psa = psa_obj)
head(evpi)
plot(evpi)
```

### Expected loss curves

**explanation here**

```{r}
el <- calc_exp_loss(wtp = example_psa$wtp, psa = psa_obj)
head(el)
plot(el, n_x_ticks = 8, n_y_ticks = 6)
```

## One- and Two-Way Sensitivity Analyses

### One-Way

Make an `owsa` object.

```{r}
o <- owsa(psa_obj)
```

Default plot:

```{r fig.height = 6, fig.width = 6}
plot(o)
```

Tornado plot:

```{r}
owsa_tornado(o, strategy = "Radio")
```

Can filter out strategies that don't change using `min_rel_diff`.

```{r}
owsa_tornado(o, strategy = "Radio", min_rel_diff = 0.05)
```

Can also return plot data:

```{r}
owsa_tornado(o, strategy = "Radio", return = "data")
```

Optimal strategy rangeplot:

```{r, fig.height = 6, fig.width = 8}
owsa_opt_strat(o)
```

Plot data:

```{r}
owsa_opt_strat(o, return = "data")
```

### Two-way

specify the two parameters of interest.

```{r}
tw <- twsa(psa_obj, parm1 = "pFailChemo", parm2 = "muDieCancer")
```

Plot

```{r}
plot(tw)
```

## Base-case CEA

Summarize the PSA results:

```{r}
psa_sum <- summary(psa_obj)
```

Calculate incremental cost-effectiveness ratios:

```{r fig.height = 5, fig.width = 7}
icers <- calculate_icers(cost = psa_sum$meanCost, effect = psa_sum$meanEffect, strategies = psa_sum$Strategy)
plot(icers)
```

Label more:

```{r fig.height = 5, fig.width = 7}
plot(icers, label = "all")
```
