---
title: "Probabilistic Sensitivity Analysis: Generation"
author: "Fernando Alarid-Escudero, Greg Knowlton, Eva Enns, and the DARTH Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Probabilistic Sensitivity Analysis: Generation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```

# Overview

This vignette explains how to use `dampack` to generate your own PSA using only a decision analytic model and information about the distributions that define your model parameters of interest. If both costs and effects are calculated within the decision model, the resulting `psa` object will be compatible with all of `dampack`'s PSA analysis functions, which are explained at length in the `psa_analysis` vignette (type `vignette("psa_generation", package = "dampack")` in the console to view this vignette).

## Decision Model Format

In order to generate a PSA in `dampack`, the user must input the code for the decision analytic model in a standardized format that is compatible with the `run_psa` function. This is the same format required for the `FUN` argument of `run_owsa_det` and `run_twsa_det`.

The user-defined model function must accept a single input containing a list of the parameters from the `params_basecase` argument. In the example model shown below, this list is named `l_params_all`, and the variables contained in this list are the only variables that are allowed to be varied through the `params_range` argument in the DSA. Optionally, additional function inputs for `FUN` can be supplied through the `...` argument of `run_owsa_det`/`run_twsa_det`, but these inputs are not allowed to vary in the sensitivity analysis. These additional inputs must be arguments of `FUN`, like `n_wtp` in the example of `calculate_ce_out()` below. `FUN` and its component functions are also able to incorporate variables stored in the global environment, such as `n_age_init` or `n_age_max` in the example.

The user-defined model function must return a data.frame where the first column contains a character vector of the strategy names, and the subsequent columns contain numeric vectors of all relevant model outcomes. Each row of the data.frame will consist of a strategy name followed by the corresponding outcome values for that strategy. These model outcomes must be calculated internally within `FUN`. The model outcomes are not limited to typical outcomes like cost or effectiveness and can be any numerical outcome that the user chooses to model.

```{r}
library(dampack)
v_names_str <- c("No Treatment", "Treatment")  
## Number of strategies
n_str <- length(v_names_str)
## Markov model parameters
n_age_init <- 25                # age at baseline
n_age_max  <- 55                # maximum age of follow up
n_t  <- n_age_max - n_age_init  # time horizon, number of cycles
v_n  <- c("H", "S1", "S2", "D")               # the 4 states of the model: Healthy (H), Sick (S1), Sicker (S2), Dead (D)
n_s <- length(v_n)                            # number of health states 

calculate_ce_out <- function(l_params_all, n_wtp = 100000){
  # User defined
  with(as.list(l_params_all), {
    ## Create discounting vectors
    v_dwc <- 1 / ((1 + d_e) ^ (0:(n_t))) # vector with discount weights for costs
    v_dwe <- 1 / ((1 + d_c) ^ (0:(n_t))) # vector with discount weights for QALYs
    
    ## Run STM model at a parameter set for each intervention
    l_model_out_no_trt <- decision_model(l_params_all = l_params_all)
    l_model_out_trt    <- decision_model(l_params_all = l_params_all)
    
    ## Cohort trace by treatment
    m_M_no_trt <- l_model_out_no_trt$m_M # No treatment
    m_M_trt    <- l_model_out_trt$m_M    # Treatment
    
    ## Vectors with costs and utilities by treatment
    v_u_no_trt <- c(u_H, u_S1, u_S2, u_D)
    v_u_trt    <- c(u_H, u_Trt, u_S2, u_D)
    
    v_c_no_trt <- c(c_H, c_S1, c_S2, c_D)
    v_c_trt    <- c(c_H, c_S1 + c_Trt, c_S2 + c_Trt, c_D)
    
    ## Mean Costs and QALYs for Treatment and NO Treatment
    v_tu_no_trt <- m_M_no_trt %*% v_u_no_trt
    v_tu_trt    <- m_M_trt %*% v_u_trt
    
    v_tc_no_trt <- m_M_no_trt %*% v_c_no_trt
    v_tc_trt    <- m_M_trt %*% v_c_trt
    
    ## Total discounted mean Costs and QALYs
    tu_d_no_trt <- t(v_tu_no_trt) %*% v_dwe 
    tu_d_trt    <- t(v_tu_trt) %*% v_dwe
    
    tc_d_no_trt <- t(v_tc_no_trt) %*% v_dwc
    tc_d_trt    <- t(v_tc_trt)    %*% v_dwc
    
    ## Vector with total discounted mean Costs and QALYs
    v_tc_d <- c(tc_d_no_trt, tc_d_trt)
    v_tu_d <- c(tu_d_no_trt, tu_d_trt)
    
    ## Vector with discounted net monetary benefits (NMB)
    v_nmb_d <- v_tu_d * n_wtp - v_tc_d
    
    ## Dataframe with discounted costs, effectiveness and NMB
    df_ce <- data.frame(Strategy = v_names_str,
                        Cost     = v_tc_d,
                        Effect   = v_tu_d,
                        NMB      = v_nmb_d)
    
    return(df_ce)
  }
  )
}

decision_model <- function(l_params_all, verbose = FALSE) {
  with(as.list(l_params_all), {
    # compute internal parameters as a function of external parameters
    r_HD    = - log(1 - p_HD) # rate of death in healthy
    r_S1D   = hr_S1 * r_HD 	  # rate of death in sick
    r_S2D   = hr_S2 * r_HD  	# rate of death in sicker
    p_S1D   = 1 - exp(-r_S1D) # probability to die in sick
    p_S2D   = 1 - exp(-r_S2D) # probability to die in sicker
    
    ####### INITIALIZATION ##########################################
    # create the cohort trace
    m_M <- matrix(NA, nrow = n_t + 1 , 
                  ncol = n_s,
                  dimnames = list(0:n_t, v_n)) # create Markov trace 
    
    m_M[1, ] <- c(1, 0, 0, 0)                  # initialize Markov trace
    
    # create transition probability matrix for NO treatment
    m_P <- matrix(0,
                  nrow = n_s, 
                  ncol = n_s,
                  dimnames = list(v_n, v_n))
    # fill in the transition probability array
    ### From Healthy
    m_P["H", "H"]  <- 1 - (p_HS1 + p_HD)
    m_P["H", "S1"] <- p_HS1
    m_P["H", "D"]  <- p_HD
    ### From Sick
    m_P["S1", "H"]  <- p_S1H
    m_P["S1", "S1"] <- 1 - (p_S1H + p_S1S2 + p_S1D)
    m_P["S1", "S2"] <- p_S1S2
    m_P["S1", "D"]  <- p_S1D
    ### From Sicker
    m_P["S2", "S2"] <- 1 - p_S2D
    m_P["S2", "D"]  <- p_S2D
    ### From Dead
    m_P["D", "D"] <- 1
    
    # check rows add up to 1
    if (!isTRUE(all.equal(as.numeric(rowSums(m_P)), as.numeric(rep(1, n_s))))) {
      stop("This is not a valid transition Matrix")
    }
    
    ############# PROCESS ###########################################
    
    for (t in 1:n_t){   # throughout the number of cycles
      m_M[t + 1, ] <- m_M[t, ] %*% m_P  # estimate the Markov trace for cycle the next cycle (t + 1)
    }
    
    ####### EPIDEMIOLOGICAL OUTPUT  ###########################################
    #### Overall Survival (OS) ####
    v_os <- 1 - m_M[, "D"]  # calculate the overall survival (OS) probability for no treatment
    
    #### Disease prevalence #####
    v_prev <- rowSums(m_M[, c("S1", "S2")])/v_os
    
    #### Proportion of sick in S1 state #####
    v_prop_S1 <- m_M[, "S1"] / v_prev
    
    ####### RETURN OUTPUT  ###########################################
    out <- list(m_M = m_M,
                m_P = m_P,
                Surv = v_os[-1],
                Prev = v_prev[-1],
                PropSick = v_prop_S1[c(11, 21, 31)])
    
    return(out)
  }
  )
}
```

## Generating Parameter Samples for a PSA

The `gen_psa_samp` function creates a `data.frame` of parameter value samples based on the underlying distributions specified by the user. Each row of the returned `data.frame` is an independently sampled set of the parameters varied in the PSA. To produce a `psa` object, the `run_psa` function will take each row of this `data.frame` and calculate the outcomes for each strategy in the user-defined model. The `data.frame` returned by `gen_psa_samp` matches the format required by the `parameters` argument of the `make_psa_obj` function.

`gen_psa_samp` has five arguments: `params` is a vector containing the names of each parameter to be varied in the PSA; `dists` is a vector of the same length indicating which type of distribution this parameter will be drawn from; `parameterization_types` is a vector indicating the format of how these parameter distributions are defined; `dists_params` is a list of vectors, where each element of the list contains the values necessary to define the distribution for a parameter based upon its corresponding element of `dists` and `parameterization_types`; and finally, `nsamp` is a numeric value indicating the number of PSA samples to be generated. 

Details about the allowable distributions, their parameterization types and the corresponding formats for `dists_params` can be found in the help documentation by typing `?gen_psa_samp` in the console. Within the example below, the first parameter in the PSA, `"p_HS1"`, follows a `"beta"` distribution, which has an `"a, b"` parameterization type (which stands for alpha, beta), and the two values for alpha and beta are `30` and `170`, respectively. If the user does not possess estimates for the alpha and beta parameters for the beta distribution but does have estimates for the mean and standard deviation of `"p_HS1"`, they also could choose to parameterize this distribution using `parameterization_types = "mean, sd"`. In this case, the first element of the dists_params list would need to be a numeric vector of length 2 containing the estimated mean and standard deviation for `"p_HS1"`. `dampack` would then use a method-of-moments estimator to calculate an alpha and beta parameter for this distribution from which the PSA sample values are drawn.

```{r}
my_params <- c("p_HS1", 
               "p_S1H", 
               "p_S1S2", 
               "p_HD", 
               "hr_S1", 
               "hr_S2", 
               "c_H", 
               "c_S1", 
               "c_S2", 
               "c_Trt",
               "u_H",
               "u_S1",
               "u_S2", 
               "u_Trt")

my_dists <- c("beta", 
              "beta", 
              "beta",
              "beta", 
              "log-normal", 
              "log-normal",
              "gamma", 
              "gamma", 
              "gamma",
              "gamma", 
              "truncated-normal", 
              "truncated-normal", 
              "truncated-normal", 
              "truncated-normal")

my_parameterization_types <- c("a, b", 
                               "a, b", 
                               "a, b",
                               "a, b", 
                               "meanlog, sdlog", 
                               "meanlog, sdlog",
                               "shape, scale", 
                               "shape, scale", 
                               "shape, scale",
                               "shape, scale", 
                               "mean, sd, ll, ul", 
                               "mean, sd, ll, ul", 
                               "mean, sd, ll, ul", 
                               "mean, sd, ll, ul")

my_dists_params <- list(c(30, 170), 
                        c(60, 60), 
                        c(84, 716),
                        c(10, 1990), 
                        c(log(3), 0.01), 
                        c(log(10), 0.02),
                        c(100, 20), 
                        c(177.8, 22.5), 
                        c(225, 66.7),
                        c(73.5, 163.3), 
                        c(1, 0.01, NA, 1),
                        c(0.75, 0.02, NA, 1), 
                        c(0.5, 0.03, NA, 1),
                        c(0.95, 0.02, NA, 1))
                       
my_psa_params <- gen_psa_samp(params = my_params,
                              dists = my_dists,
                              parameterization_types = my_parameterization_types,
                              dists_params = my_dists_params,
                              n = 100)
```

## Generating Outcomes for the PSA

The `run_psa` function is used to calculate outcomes for each strategy for every PSA sample through the user-defined decision model, `FUN`. In this example, the `data.frame` of PSA parameters generated by `gen_psa_samp` should be used as the input for the `psa_samp` argument. The combination of the parameters in the `psa_samp` and the `params_basecase` argument must define every parameter that `FUN` expects within its `l_params_all` input argument. Other parameters for `FUN` that are not contained within `l_params_all` list, like the `n_wtp` argument of `calculate_ce_out` can be passed through `...` as an additional argument in `run_psa`. If the decision model in `FUN` is computationally slow and/or the number of PSA samples is extremely large, `run_psa` could take a long time to run. Under these circumstances, it is recommended that you set the `progress` argument to `TRUE` in order to print a progress bar in the console while the function is running to monitor its progress.

```{r}
l_params_basecase <- list(u_D = 0, 
                          c_D = 0, 
                          d_e = 0.03, 
                          d_c = 0.03,
                          p_HS1 = NA, 
                          p_S1H = NA, 
                          p_S1S2 = NA, 
                          p_HD = NA,
                          hr_S1 = NA, 
                          hr_S2 = NA, 
                          c_H = NA, 
                          c_S1 = NA,
                          c_S2 = NA, 
                          c_Trt = NA, 
                          u_H = NA, 
                          u_S1 = NA, 
                          u_S2 = NA,
                          u_Trt = NA)

psa_output <- run_psa(psa_samp = my_psa_params,
                      params_basecase = l_params_basecase,
                      FUN = calculate_ce_out,
                      outcomes = c("Cost", "Effect", "NMB"),
                      strategies = c("No_Treatment", "Treatment"),
                      n_wtp = 150000,
                      progress = FALSE)
```

## Creating a Fully-functional PSA Object

`run_psa` will return a named list containing a `psa` object for each outcome specified in the `outcomes` argument. Each `psa` object in the list is compatible with `owsa()`, `twsa()`, and their associated downstream functions described in the `psa_analysis` vignette. However, most PSA analysis functions in `dampack` rely on the clear designation of both a cost and effectiveness outcome. To create a PSA object that is compatible with these functions related to cost-effectiveness you must input the results of `run_psa` into the `make_psa_obj` function in the following manner. `make_psa_obj()` requires data.frames for `cost`, `effect`, and `parameters`, and a character vector for `strategies`. The `data.frame`s containing each outcome in the list returned by `run_psa` are stored within `other_outcome`. In this example, the outcome associated with `effect` is named `"Effect"`, and so `psa_output$Effect$other_outcome` is supplied to the corresponding argument of `make_psa_obj`.

```{r}
cea_psa <- make_psa_obj(cost = psa_output$Cost$other_outcome, 
                        effect = psa_output$Effect$other_outcome, 
                        parameters = psa_output$Cost$parameters,
                        strategies = psa_output$Cost$strategies,
                        currency = "$")

```


